---
title: "Age-Standardised Testicular Cancer Incidence In The Nordic Countries"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include = FALSE}
# knitting runs from the project root, like 01 and 02. root.dir does not apply
# when chunks are run one at a time in RStudio, so paths go through proj_file()
knitr::opts_knit$set(root.dir = "..")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 150)
```

Data: **NORDCAN 2.0**, Association of the Nordic Cancer Registries / IARC,
<https://nordcan.iarc.fr>, version 9.6 (06.2026), exported 20 September 2026.
The exploratory report carries the full source description.

```{r shared-setup}
source(if (file.exists("R/00_setup.R")) "R/00_setup.R" else "00_setup.R")
```

```{r libraries}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PHEindicatormethods)
library(broom)
```

# Load

```{r load}
age_specific <- read_csv(proj_file("data/processed/testis_age_specific.csv"),
                         show_col_types = FALSE)

published <- read_csv(proj_file("data/processed/nordcan_published.csv"),
                      show_col_types = FALSE)

# stored as text, so the default sort would run 0-4, 10-14, 15-19, ..., 5-9, 85+
band_levels <- age_specific |>
  distinct(age_start, age_band) |>
  arrange(age_start) |>
  pull(age_band)

# the weight vector is positional, so row order inside each group is
# load-bearing. Checked as the file arrived, before any sorting of our own
rows_arrived_sorted <- all(tapply(age_specific$age_start,
                                  paste(age_specific$country, age_specific$period),
                                  \(x) !is.unsorted(x)))

age_specific <- age_specific |>
  mutate(age_band = factor(age_band, levels = band_levels, ordered = TRUE)) |>
  arrange(country, period, age_start)

band_order <- age_specific |>
  distinct(age_band, age_start) |>
  arrange(age_band)

stopifnot(
  "age_band levels are not the 18 expected five-year bands" =
    identical(levels(age_specific$age_band), EXPECTED_BANDS),
  "age_band level order does not ascend by age" =
    all(diff(band_order$age_start) > 0),
  "age_band is not an ordered factor" =
    is.ordered(age_specific$age_band),
  "age_specific is not countries x periods x age bands" =
    nrow(age_specific) == N_COUNTRIES * N_PERIODS * N_BANDS,
  "published is not countries x periods" =
    nrow(published) == N_COUNTRIES * N_PERIODS,
  "a country-period is missing some of its 18 age bands" =
    all(table(age_specific$country, age_specific$period) == N_BANDS),
  "the same age band appears twice in a country-period" =
    anyDuplicated(age_specific[c("country", "period", "age_start")]) == 0,
  "missing values in the prepared data" =
    !anyNA(age_specific) && !anyNA(published),
  "the input arrived with age bands out of order inside a country-period" =
    rows_arrived_sorted,
  "period_mid is not the midpoint of its own period" =
    all(age_specific$period_mid ==
          as.integer(substr(age_specific$period, 1, 4)) + 2)
)
```

These are the same two files the exploratory analysis used, and I run the same
checks over them again. I would rather not assume the earlier stage left
everything tidy.

# The Standard Population

Countries cannot be compared on raw rates, because their age profiles differ and
the exploratory stage showed those profiles are shifting over time as well. The
way round it is to reweight every country onto one shared reference population,
so each is measured as if its age structure were the same as everyone else's.
That reference is the **European Standard Population 2013**, which is also what
NORDCAN uses for its own published figures, so our numbers and theirs stay
directly comparable.

One thing has to be squared away first. The published ESP2013 splits the
population into 19 age groups and runs up to "90 and over", while this data
stops at "85 and over". The two have to line up before any reweighting can
happen, so the top two ESP2013 groups are added together into a single "85 and
over" group, which leaves 18 to match the data. The weights are read straight
from the package rather than typed in by hand, so a mistyped number cannot
quietly change every result downstream.

```{r standard-population}
data(esp2013, package = "PHEindicatormethods")

esp18 <- c(esp2013[1:17], esp2013[18] + esp2013[19])

stopifnot(
  "the published ESP2013 vector is not 19 bands" = length(esp2013) == 19,
  "the collapsed vector is not 18 bands" = length(esp18) == N_BANDS,
  "collapsing changed the total weight" = sum(esp18) == sum(esp2013),
  "the top band is not the sum of the last two published weights" =
    esp18[18] == esp2013[18] + esp2013[19]
)

tibble(Measure = c("Published Bands", "Collapsed Bands",
                   "Total Weight", "Top Band Weight"),
       Value   = c(length(esp2013), length(esp18), sum(esp18), esp18[[18]]))
```

The table confirms the fold did what it was meant to. Nineteen groups became
eighteen, the top group now carries 2,500, which is the two old top weights
added together, and the total is still 100,000.

# Calculating The Standardised Rates

Everything that follows rests on this one calculation, so I work it out three
times and check the answers against each other rather than trusting any single
one. Once by hand, straight from the formula. Then with `phe_dsr()`, a
function from a maintained public health
package. Then with `calculate_dsr()`, the newer function that replaces it. If
all three land on the same number, the calculation is doing what it is supposed
to.

The two package functions are separate pieces of code but come from the same
package and use the same underlying method, so this is really one hand
calculation checked against two implementations rather than three fully
independent routes.

```{r asr-three-ways}
# the formula itself: each band's rate, weighted by the standard population
asr_hand <- age_specific |>
  group_by(country, period, period_mid) |>
  summarise(asr_hand = sum(esp18 * cases / person_years) / sum(esp18) * 1e5,
            .groups = "drop")

# phe_dsr takes the weights as a vector
asr_phe95 <- age_specific |>
  group_by(country, period) |>
  phe_dsr(cases, person_years, stdpop = esp18, stdpoptype = "vector",
          confidence = 0.95) |>
  select(country, period, asr = value, lower95 = lowercl, upper95 = uppercl)

# two levels in one call are only allowed as 0.95 and 0.998, so 99% runs separately
asr_phe99 <- age_specific |>
  group_by(country, period) |>
  phe_dsr(cases, person_years, stdpop = esp18, stdpoptype = "vector",
          confidence = 0.99) |>
  select(country, period, lower99 = lowercl, upper99 = uppercl)

# calculate_dsr rejects a bare vector and wants the weights as a column
asr_calc <- age_specific |>
  group_by(country, period) |>
  mutate(stdpop = esp18) |>
  calculate_dsr(cases, person_years, stdpop = stdpop, confidence = 0.95) |>
  ungroup() |>
  select(country, period, asr_calc = value)

asr <- asr_hand |>
  left_join(asr_phe95, by = c("country", "period")) |>
  left_join(asr_phe99, by = c("country", "period")) |>
  left_join(asr_calc,  by = c("country", "period")) |>
  left_join(select(published, country, period, asr_nordcan = asr_eu2013),
            by = c("country", "period"))
```

```{r shape-after-join}
stopifnot(
  "the ASR table is not one row per country-period" =
    nrow(asr) == N_COUNTRIES * N_PERIODS,
  "a country-period lost or gained a row in the joins" =
    anyDuplicated(asr[c("country", "period")]) == 0
)
```

```{r three-way-table}
asr |>
  summarise(`Hand Against phe_dsr`        = max(abs(asr_hand - asr)),
            `phe_dsr Against calculate_dsr` = max(abs(asr - asr_calc)),
            `Ours Against NORDCAN`        = max(abs(asr - asr_nordcan))) |>
  pivot_longer(everything(), names_to = "Comparison",
               values_to = "Largest Difference")
```

All three land on the same number. Against NORDCAN's own published figure the
largest gap anywhere in the 35 country-periods is about 0.00015, which is small
enough to be the rounding they apply when publishing rather than any real
difference in how the rate was worked out.

# Validation Against NORDCAN

The section above showed the three calculations agree with each other. This one
is the formal check against an outside answer. I set two tolerances before
running any of this, and I test them here as code rather than reporting them as
prose, so the report refuses to build if either one is missed.

```{r validation}
stopifnot(
  "hand-computed ASR and phe_dsr disagree by more than rounding" =
    max(abs(asr$asr_hand - asr$asr)) < 1e-9,
  "phe_dsr and calculate_dsr disagree" =
    max(abs(asr$asr - asr$asr_calc)) < 1e-9,
  "computed ASR is outside the pre-registered 0.05 tolerance against NORDCAN" =
    max(abs(asr$asr - asr$asr_nordcan)) < 0.05
)

asr |>
  filter(period == "1990-1994") |>
  transmute(Country = country,
            `Hand Computed` = round(asr_hand, 5),
            `phe_dsr` = round(asr, 5),
            `calculate_dsr` = round(asr_calc, 5),
            `NORDCAN Published` = round(asr_nordcan, 5))
```

The table takes one period and puts all four figures side by side, so the
agreement can be seen rather than just asserted. Denmark's 1990 to 1994 rate
comes out at 10.04845 by hand, 10.04845 from each package function, and
10.04845 in NORDCAN's published table.

# The Standardised Trend

With age differences removed, the countries can finally be put on one chart and
read against each other.

```{r asr-trend}
# the five countries share an x position, so everything is nudged apart by the
# same amount to keep each country's interval readable
dodge <- position_dodge(width = 0.45)

ggplot(asr, aes(period, asr, group = country, colour = country)) +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0,
                linewidth = 0.7, alpha = 0.55, position = dodge) +
  geom_line(linewidth = 0.8, position = dodge) +
  geom_point(size = 2, position = dodge) +
  scale_colour_manual(values = COUNTRY_COLORS) +
  # headroom at the top so the legend sits clear of every line
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0.02, 0.30))) +
  labs(title = "Age-Standardised Incidence Rate By Period, With 95% Intervals",
       x = NULL, y = "ASR per 100,000 (European Standard 2013)",
       caption = SOURCE_NOTE) +
  theme_eda() +
  legend_corner(0.01, 0.99) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

What stands out is where the movement is, not that there is movement. Denmark
barely shifts across the whole window, up about 5%, while Finland more than
doubles. The countries converge because the bottom rises, not because the top
comes down, which is a different story from the one a shrinking gap might
suggest on its own. Iceland is the only country to finish lower than it started,
and across the window its interval overlaps three of the other four at one point
or another, more than any other country manages, so its place in the ranking is
the least settled of the five.

```{r asr-change}
first_last <- range(asr$period)

asr |>
  filter(period %in% first_last) |>
  select(country, period, asr) |>
  pivot_wider(names_from = period, values_from = asr) |>
  mutate(Change = .data[[first_last[2]]] - .data[[first_last[1]]],
         `Percent Change` = (.data[[first_last[2]]] /
                               .data[[first_last[1]]] - 1) * 100) |>
  arrange(desc(`Percent Change`)) |>
  rename(Country = country) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

```{r asr-table}
asr |>
  transmute(Country = country, Period = period, ASR = round(asr, 2),
            `95% Lower` = round(lower95, 2), `95% Upper` = round(upper95, 2),
            `99% Lower` = round(lower99, 2), `99% Upper` = round(upper99, 2)) |>
  arrange(Country, Period)
```

The 99% columns are a stricter version of the same interval, there as a check
rather than as the headline figure. The reasoning is in the rate ratio section
below.

The interval columns are the ones worth reading here, more than the rates
themselves. In the final period Sweden's 95% interval spans 0.64 and Iceland's
spans 3.05, nearly five times wider, for two countries whose point estimates sit
about one apart. Widening from 95% to 99% barely moves the large countries and
stretches Iceland noticeably further, which is the small-numbers problem turning
up again in a second place.

# Rate Ratio By Period

This section reduces five separate rates down to the one number the research
question actually asks for: how big is the gap.

The measure is straightforward. In each period, take the country with the
highest rate and divide it by the country with the lowest. A value of 1 would
mean no gap at all, and the further above 1 it sits the wider the spread. The
pair is not decided in advance, because the extremes really do move. Denmark and
Norway take turns at the top, Finland and Iceland take turns at the bottom, and
four different pairings show up across the seven periods.

Each ratio carries its own margin of error, worked out by combining the
uncertainty of whichever two countries are involved that period. One note for
anyone checking the arithmetic: the intervals coming out of the standardisation
step are slightly lopsided, and I treat them as even. On
this data the two halves of an interval differ by about 1.5% in a typical case
once you look at them on the log scale, so the approximation changes very
little.

```{r rate-ratio}
Z95 <- qnorm(0.975)
Z99 <- qnorm(0.995)

# a log-scale variance recovered from each country's own interval
asr <- asr |>
  mutate(var_log = ((log(upper95) - log(lower95)) / (2 * Z95))^2)

rate_ratio <- asr |>
  group_by(period, period_mid) |>
  summarise(highest_country = country[which.max(asr)],
            lowest_country  = country[which.min(asr)],
            highest_asr     = max(asr),
            lowest_asr      = min(asr),
            var_log_ratio   = var_log[which.max(asr)] + var_log[which.min(asr)],
            # do the two extremes' own intervals overlap this period
            cis_overlap     = lower95[which.max(asr)] <= upper95[which.min(asr)],
            .groups = "drop") |>
  mutate(rate_ratio = highest_asr / lowest_asr,
         se_log     = sqrt(var_log_ratio),
         lower95    = exp(log(rate_ratio) - Z95 * se_log),
         upper95    = exp(log(rate_ratio) + Z95 * se_log),
         lower99    = exp(log(rate_ratio) - Z99 * se_log),
         upper99    = exp(log(rate_ratio) + Z99 * se_log))
```

```{r rate-ratio-plot}
ggplot(rate_ratio, aes(period, rate_ratio)) +
  # the 99% bar is the same hue faded, so it reads as the wider version of the
  # same interval rather than as a second measurement
  geom_errorbar(aes(ymin = lower99, ymax = upper99), width = 0,
                linewidth = 0.9, colour = COUNTRY_COLORS[["Denmark"]],
                alpha = 0.35) +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0,
                linewidth = 1.8, colour = COUNTRY_COLORS[["Denmark"]]) +
  geom_point(size = 2.8, colour = COUNTRY_COLORS[["Denmark"]]) +
  geom_hline(yintercept = 1, linewidth = 0.3, linetype = "dashed",
             colour = INK_MUTED) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0.02, 0.08))) +
  labs(title = "Highest To Lowest Rate Ratio, With 95% And 99% Intervals",
       x = NULL, y = "Rate Ratio",
       caption = SOURCE_NOTE) +
  theme_eda() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The decline is gradual rather than sudden, and that matters for how it should be
read: a gap that narrows in almost every period is a different thing from one
that collapses once and then holds. The run of falls breaks only at the end,
where the final period edges back up from 1.79 to 1.83. The two widest intervals
on the chart, 2010 to 2014 and 2020 to 2024, are exactly the periods where
Iceland is one of the extremes, so that closing uptick is also among the least
precisely measured points on the chart.

The 99% bar is a stricter check rather than a correction. Read across seven
periods, the chance that at least one 95% interval misses the true value is
about 30%, and 99% is close to what a formal adjustment for seven comparisons
would give. Neither interval allows for choosing the highest and lowest country
each period.

```{r rate-ratio-table}
rate_ratio |>
  transmute(Period = period,
            Highest = highest_country, Lowest = lowest_country,
            `Rate Ratio` = round(rate_ratio, 2),
            `95% Interval` = sprintf("%.2f to %.2f", lower95, upper95),
            `99% Interval` = sprintf("%.2f to %.2f", lower99, upper99),
            `Extremes Overlap` = ifelse(cis_overlap, "Yes", "No"))
```

The Lowest column is the one to watch. Finland holds the bottom for the first
four periods, and from 2010 onward it no longer does so reliably, because by
then Finland and Iceland are running within about 0.1 of each other. Which of
the two comes out lower flips with Iceland's swings from period to period rather
than with any settled difference between them.

That points at something the Extremes Overlap column cannot tell you. It
compares the top country's lower bound against the bottom country's upper bound,
and those are far apart in every period, which is why it reads No throughout. It
confirms that a gap exists; it says nothing about *which* country sits at each
end. On that question the labels are shakier than they look. In 2015 to 2019
Denmark and Norway are 0.03 apart with almost identical intervals, so naming
Denmark the highest is close to a coin flip, and the same goes for Finland
against Iceland at the bottom in 2010 to 2014 and 2020 to 2024.

The reassuring part is that this barely touches the answer. Swapping every one
of those near-tied labels moves the ratios by at most 0.035 and the trend slope
from -0.0291 to -0.0293. The size of the gap is solid. The names attached to its
two ends, in those particular periods, are not.

# Trend Test

Looking at the chart above, the gap obviously shrinks. I did not want to rest on
reading a slope off a picture though, so I tested it properly. I fitted a
straight line through the seven ratios against the midpoint year of each period,
working on the log scale so the slope reads as a percentage change rather than a
fixed amount, and I weighted each period by how certain it is, so the shakier
readings pull the line around less than the solid ones.

```{r trend-test}
trend_fit <- lm(log(rate_ratio) ~ period_mid,
                data = rate_ratio, weights = 1 / var_log_ratio)

trend_tidy <- tidy(trend_fit, conf.int = TRUE)

trend_tidy |>
  filter(term == "period_mid") |>
  transmute(`Slope Per Year` = round(estimate, 5),
            `Standard Error` = round(std.error, 5),
            `95% Lower` = round(conf.low, 5),
            `95% Upper` = round(conf.high, 5),
            `P Value` = signif(p.value, 3),
            `Annual Change` = sprintf("%.2f%%", (exp(estimate) - 1) * 100),
            `Change Over 30 Years` = sprintf("%.1f%%", (exp(estimate * 30) - 1) * 100))
```

The slope comes out around twenty times its own standard error, which is why the
p value lands where it does. A decline this consistent is not the sort of thing
seven noisy points throw up by chance. A more intuitive way to read the annual
figure is this: keep going at that rate and the gap halves roughly every 24
years.

One thing worth holding onto, though. This measures the trend in the *ratio*,
not in anyone's risk. A ratio can shrink while every country's rate is climbing,
and that is close to what is happening here.

```{r trend-se}
tq    <- qt(0.975, df.residual(trend_fit))
slope <- coef(trend_fit)[["period_mid"]]

se_relative <- sqrt(diag(vcov(trend_fit)))[["period_mid"]]
se_exact    <- sqrt(diag(vcov(trend_fit)) /
                      summary(trend_fit)$sigma^2)[["period_mid"]]

tibble(Weighting = c("Weights Treated As Relative",
                     "Weights Treated As Exact"),
       se = c(se_relative, se_exact)) |>
  transmute(Weighting,
            `Standard Error` = round(se, 5),
            `95% Lower` = round(slope - tq * se, 4),
            `95% Upper` = round(slope + tq * se, 4),
            `Fall Over 30 Years` =
              sprintf("%.0f%% to %.0f%%",
                      -(exp((slope + tq * se) * 30) - 1) * 100,
                      -(exp((slope - tq * se) * 30) - 1) * 100))
```

The weights are treated as relative rather than exact, which is how the fitting
function handles them by default, so the standard error reflects how closely the
seven points sit around the line. They sit closer than their stated
uncertainties implied. Taking those uncertainties at face value instead widens
the 95% interval on the slope from -0.0328 to -0.0254 out to -0.0356 to -0.0226.
That is a fall in the gap of roughly 49% to 66% over the 30 years, instead of
53% to 63%. The conclusion is the same either way.

```{r trend-plot}
ggplot(rate_ratio, aes(period_mid, log(rate_ratio))) +
  geom_smooth(aes(weight = 1 / var_log_ratio), method = "lm", formula = y ~ x,
              colour = COUNTRY_COLORS[["Denmark"]], fill = GRID, linewidth = 0.8) +
  geom_point(aes(size = 1 / var_log_ratio),
             colour = COUNTRY_COLORS[["Denmark"]], show.legend = FALSE) +
  scale_size_continuous(range = c(1.8, 5)) +
  scale_x_continuous(breaks = rate_ratio$period_mid) +
  labs(title = "Weighted Trend In The Log Rate Ratio",
       x = "Period Midpoint Year", y = "Log Rate Ratio",
       caption = SOURCE_NOTE) +
  theme_eda()
```

The slope is negative and clearly separated from zero, at roughly 2.9% off the
ratio each year with a p value near 0.0000056. Carried across the 30 years
between the first and last midpoints, that is a fall of about 58% in the size of
the gap. Larger points on the chart are the periods with tighter intervals,
which is the weighting made visible: the later Iceland-driven periods count for
less in fitting the line.

The gap between the highest and lowest Nordic country has narrowed substantially
across the study window, and the narrowing is larger than the uncertainty around
it. Why it narrowed is a separate question that this project does not take up.

```{r write-output, include = FALSE}
dir.create(proj_file("output"), showWarnings = FALSE)

asr_out <- asr |>
  select(country, period, period_mid, asr,
         lower95, upper95, lower99, upper99) |>
  arrange(country, period)

ratio_out <- rate_ratio |>
  select(period, period_mid, highest_country, lowest_country,
         highest_asr, lowest_asr, rate_ratio,
         lower95, upper95, lower99, upper99, cis_overlap)

trend_out <- trend_tidy |>
  transmute(term, estimate, std_error = std.error, statistic,
            p_value = p.value, conf_low = conf.low, conf_high = conf.high)

write_csv(asr_out,   proj_file("output/asr_by_country_period.csv"))
write_csv(ratio_out, proj_file("output/rate_ratio_by_period.csv"))
write_csv(trend_out, proj_file("output/trend_test.csv"))
```
